---
title: "2024-02-12 - More figures, tables, and summary stats in R"
author: "Austin K Lien"
date: "2024-02-13"
output: 
  html_document:
    # code_folding: hide
    df_print: "kable"
    toc: yes
    toc_float: yes
    toc_depth: 4
    toc_width: 1
    fig_retina: 2
    number_sections: no
    highlight: pygments
    theme: sandstone
    
  pdf_document:
    toc: yes
    toc_depth: '4'
---

# Picking up where we left off...

**During the previous two sessions we covered the follow**
- basics of the R environment and syntax in R
- setting the working directory
- bringing in data
  - factoring distinct groups
  - pivot_longer() data manipulation
- data visualization using ggplot2 and ggpubr

- Below, I will bring in all of data so we can start fresh here. 

## loading packages
```{r}
library(readxl)
library(tidyverse)
library(ggpubr)
library(RColorBrewer)
```

## Setting up Data
```{r}
setwd("C:/Users/lienx/Documents/2024-02-06 R Workshop")

Assay <- read_excel(("2024-02-06 Data Set.xlsx"), 
                      sheet = "Soil Assays")

Assay$Rep <- as.factor(Assay$Rep)

Assay$Trmt <- factor(Assay$Trmt, 
                    levels = c('kab', 'tach'),
                    labels = c("Kabina", "Tachigaren"))

Assay$Soil <- as.factor(Assay$Soil)

Assay$Year <- ordered(Assay$Year,
                      levels = c('2024', '2023'),
                      labels = c('2024', '2023'))


Assay_long <- pivot_longer(data = Assay,
                           col = c("TotalRRI", "AphRRI", "RhizRRI"),
                           names_to = "RRI_Type",
                           values_to = "Index_Value")

Assay_long$RRI_Type <- ordered(Assay_long$RRI_Type,
                     levels = c("TotalRRI", "AphRRI", "RhizRRI"),
                     labels = c("Total Root Rot",
                                "Aphanomyces Root Rot",
                                "Rhizoctonia Root Rot"))
```

## ggpubr boxplot


Let's make a box and whisker plot  of our root rot indices and add error bars and value labels using **ggboxplot()** from the ggpubr package. 

- Take note of the minor differences when setting up the plot
  - To view all options and how to use them, use the **help tab** 
- Note that all variables are in quotations
    - This also allows us to access column names that have spaces in them or columns that start with numbers
- Once the plot is set up, we can add ggplot2 objects to it

Let's also change the color palette using the **scale_fill_brewer(palette = )** function from the RColorBrewer package. 

```{r}
ggboxplot(Assay_long,
          x = "RRI_Type",
          y = "Index_Value",
          fill = "Trmt",
          facet.by = "Soil", ncol = 2) +
  scale_fill_brewer(palette = "Dark2") +
  
  theme(axis.text.x = element_text(size = 15)) +
  
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
# sets values for discrete scale aesthetics - this staggers the x axis labels


```


## ggpubr barplot

Let's make a bar plot  of our root rot indices and add error bars and value labels using **ggbarplot()** from the ggpubr package. 

    
```{r}
ggbarplot(Assay_long,
          x = "RRI_Type",
          y = "Index_Value",
          fill = "Trmt",
          
          facet.by = c("Soil", "Year"), nrow = 2,
          # slightly different than ggplot2
          
          add = 'mean_se',
          # This adds standard error bars 
          
          position = position_dodge(0.8),
          # this separates the columns, rather than being stacked
          
          # BOTH 'add = "mean",' AND 'position = position_dodge(1)' NEED TO BE USED FOR BARPLOTS
          
          label = TRUE,
          lab.size = 4,
          lab.vjust = 0.5,
          lab.nb.digits = 1,
          # these control the data labels
          
          xlab = "Root Rot Index Type",
          ylab = "Root Rot Index (0-100)",
          title = "Root Rot Indices of disease nursery soils by year"
          # We can change these titles here, but not the legend title
          ) +
  
   labs(fill = "Seed Treatment") + 
  
  scale_fill_brewer(palette = "Dark2") +
    # This comes from the RColorBrewer Package
    
  theme(axis.text.x = element_text(size = 7.5),
        legend.position = c(0.9,0.9)) 
        # You can place the legend anywhere within the plot using a vector
    

```

## ggpubr boxplot and jitter

**Let's try a different plot with ggpubr**

- We will use **ggboxplot()** from the ggpubr package
- I am also introducing **jitter** points and manually changing shapes or colors

```{r}

ggboxplot(Assay,
          x = "Soil",
          y = "AphRRI",
          fill = "Soil",
          add = "jitter",
          add.params = list(shape = "Trmt"),
          # this is adding jitter points according to treatment
          # we could also assign the color

          ) +
scale_shape_manual(values = c(2,3)) +
# This manually changed the jitter shapes (shape = ) 
  
scale_fill_manual(values = c("salmon", "cyan"))
# This changed the boxplot (fill = ) color
```

**We can also view the distribution of the data using a violin plot**

## violin plot
```{r}
ggviolin(Assay, 
          x = "Trmt",
          add = 'jitter',
          fill = "Soil",
          y = "AphRRI",
          add.params = list(shape = "Year",
                            fill = "Soil")) +
  scale_shape_manual(values = c(2,4))
```


## ggpubr line graph

Now, let's create line graphs using **ggline()** to display trends and patterns over a continuous variable, making them suitable for time series data or any data with a natural ordering.

The columns labelled c('DAP_6', 'DAP_14', 'DAP_21', 'DAP_28') are Days After Planting (DAP).

The data within the these columns are the number of living plants expressed as a percentage of the total number of planted. The column 

**First, we need to create a new data.frame and use pivot_longer() to transpose these columns. We can also rename our values within our new column and make sure they are in the correct order**


```{r}
Assay_plants <- pivot_longer(Assay,
                             cols = c('DAP_6', 'DAP_14', 'DAP_21', 'DAP_28'),
                             names_to = "DAP",
                             values_to = "plants")

Assay_plants$DAP <- ordered(Assay_plants$DAP,
                            levels = c('DAP_6', 'DAP_14', 'DAP_21', 'DAP_28'),
                            labels = c('6','14','21','28'))
```

Now, we can use **ggline()** to create a line graph with DAP on the x axis. 

- I will also introduce the **guides()** argument to customize labels in our legend and **rremove()** to remove specific components from a plot.

```{r, warning=FALSE, message=FALSE}
ggline(Assay_plants,
       x = "DAP",
       y = "plants",
       group = "Trmt",
       color = 'Trmt',
       linetype = 'Trmt',
       shape = "Trmt",
       
       add = c('mean_se'),
       # without adding 'add = mean' or another variation, the line graph will not be logical
       
       facet.by = 'Soil',
       
       plot_type = 'b',
       # creates both lines and points
       
       point.size = 1.2,
       # this changes the size of the points
       
       ylab = "Plant Stand (%)",
       xlab = "Days after planting",
       
       numeric.x.axis = TRUE
       # makes x axis as numeric (i.e., logical spacing between values)
       ) +
  
  theme_bw() +
  
  theme(text = element_text(
                            # family = "Times New Roman", 
                            # we can use the family argument once you install and load the 'extrafont' package
                            size = 12, 
                            face = "bold",
                            color = "black"),
        axis.text.x = element_text(angle = 0, 
                                   size = 10, 
                                   color = "black", 
                                   vjust = 0.5, 
                                   hjust = 0.5),
        legend.position = "bottom",
        legend.key.width = unit(1.5, "cm")) +
        # This changes the width inside the legend, making them easier to distinguish
        
  rremove('legend.title') +
  # rremove can be used to remove a specific component from a plot
    
  guides(color = guide_legend(nrow = 2)) +
  # guides can be used to customize legends
    
  scale_color_brewer(palette = 'Set2')
```


## ggpubr scatter plot

We will now use **ggscatter()** to visualize the relationship between RhizRRI and AphRRI

```{r}
ggscatter(Assay_plants, 
          x= "RhizRRI", 
          y = "AphRRI",
          
          add = "reg.line", 
          # displays regression line
          
          conf.int = TRUE,
          # displays buffer around regression line representing confidence interval
          
          cor.coef = TRUE,
          cor.coef.size = 10,
          cor.method = "pearson",
          # displays correlation coefficient, size, and method
          
          cor.coef.coord = c(50, 75)
          # moves the location of the data displayed in the plot
          )
```

**Now, let's see if there is difference among the soils.**

- We will need to use **stat_cor()** argument for displaying the coefficients for each treatment.
  - ggscatter will display both lines but will only display one coefficient. 
  
```{r}
ggscatter(Assay_plants, 
          x= "RhizRRI", 
          y = "AphRRI",
          
          color="Trmt",
          add = "reg.line", 
          conf.int = TRUE,
          )+
        # Displays regression lines and confidence interval for each factor signified by color
  
  stat_cor(aes(color = Trmt), 
           method = "pearson", 
           label.x = 40,
           digits = 3)
  # extension of ggpubr to further customize correlation stats, here we are displaying the coefficients for each Trmt.
  

```


# Piping and Creating Tables 

Tables are essential for presenting detailed data summaries and results, especially in research and data analysis.

the package **kableExtra** simplifies the creation of aesthetically pleasing and customizable tables, enhancing the presentation of summary statistics and other tabular data.

- kable() is a function from the knitr package, and it is primarily used for converting data frames into markdown or LaTeX tables.
  - It provides basic functionality for formatting tables, such as aligning columns, setting column names, and applying basic formatting options.
  - **kableExtra** is an extension package that builds upon kable() to provide additional customization options and formatting capabilities for tables with many arguments hidden in the background
    -  this allows for more advanced customization of tables, making it suitable for creating publication-quality tables with complex formatting requirements.

```{r}
library(kableExtra)

kbl(Assay)
  # creates a table with auto formatting 
```


**But to go any further, I need to introduce piping!**

- Piping simplifies the process of performing multiple operations on data by passing the output of one function directly as the input to another function.
  - The pipe operator **%>%** is used to chain operations together. 
    - It takes the output of the expression on its left-hand side and feeds it as the first argument to the function on its right-hand side.
    - Piping makes code more readable and easier to understand, especially when performing multiple transformations on data.
    - It reduces the need for intermediate variables and allows for more compact code.
    - Piping is commonly used in data manipulation, data wrangling, and analysis workflows, where multiple operations are performed sequentially on datasets.
    - The keyboard shortcut is **Ctrl+Shft+M**

```{r}
kbl(Assay) %>% 
  kable_classic(full_width = F, html_font = "Arial")
  # uses the built-in bootstrap themes by default 
```

We can also take our dataframe and pipe directly into the kbl() function and add some additional styling

```{r}

Assay %>% 
kbl(digits = 1) %>%
  # sets number of digits after the decimal point for all numbers
  
  kable_classic(full_width = F, html_font = "Arial") %>% 
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive"))
 # provides a clean approach to modify the style of HTML tables
```



# Getting summary statistics 

Summary statistics provide valuable insights into the characteristics and distributions of the data.

**rstatix** offers a convenient way to compute various summary statistics, such as means, medians, standard deviations, and correlations, streamlining the data analysis process.

```{r}
library(rstatix)

Assay %>%
get_summary_stats(type="common")
```

**However, this table won't look to good once we knit our code, so let's use what we learned with kableExtra to make it look better**

```{r}
Assay %>%
get_summary_stats(type="common") %>% 
  # common designates the type of summary statistcs (many other options are available)
  
kbl(digits = 1) %>%
    kable_classic(full_width = F, html_font = "Arial") %>%
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive"))
```

**Looks better right?**

Now what if we want to get summary stats just for a few variables?

```{r}
Assay %>%
get_summary_stats(c(TotalRRI, AphRRI, RhizRRI), type="common")  %>% 

kbl(digits = 1) %>% kable_classic(full_width = F, html_font = "Arial")
```

But what if we want to get the summary statistics for each treatment, soil, or year?

# Introducing dplyr and statistical syntax

**dplyr** and tidyr provide a set of intuitive functions for manipulating data frames, making it ideal for tasks such as filtering, summarizing, mutating, and arranging data.
  
  - dplyr and tidyr should already be loaded with tidyverse, but you can load it separately as well.
    - **group_by()** can be used to group data by one or more variables.
      - ungroup() removes grouping.
    - **filter()** can be used to subset your data based on specific criteria.
      - subset() can also be used to subset data (from base R)
    - **rename()** will rename columns. 
    - **summarise()** is used to compute specific summary statistics within each group.
      - *I prefer using rstatix*
    - **mutate()** can create new variables based on existing ones or modify existing variables.
   - Use **arrange()** to sort data based on one or more variables.
      - Continue to use the pipe operator %>% to chain together multiple dplyr operations in a single line, promoting readability and conciseness.
  - **seperate()** turns a single character column into multiple columns 
  
  
By mastering the functions provided by dplyr, you can streamline your data analysis workflow and gain greater control over your data processing tasks

Statistical operators in R are commonly used for performing various statistical calculations, hypothesis testing, and modeling. Here's an overview of some common statistical operators and their syntax in R:

COMPARISON OPERATORS

- x == y  # Equals
- x != y  # Not equals
- x > y   # Greater than
- x < y   # Less than
- x >= y  # Greater than or equal to
- x <= y  # Less than or equal to
- a :  z  # Spans between start and end



LOGICAL OPERATORS 

- x & y   # AND
- x | y   # OR
- !x      # NOT




STATISTCAL FUNCTIONS

- mean(x)      # Mean
- sd(x)        # Standard deviation
- median(x)    # Median
- cor(x, y)    # Correlation coefficient
- quantile(x)  # Quantiles



**Okay, now let's group by Soil and get some summary statistics!**

```{r}
Assay %>%
  group_by(Soil) %>% 
  # converts to group
  
get_summary_stats(TotalRRI, type= c("mean_sd"))  %>% 
# changed stats type to show only mean and standard deviation
  
kbl(digits = 1) %>% kable_classic(full_width = F, html_font = "Arial")
```

**We can also group by both Soil AND Trmt and look at only the mean and standard error Root rot indices**

```{r}
Assay %>%
  group_by(Soil:Trmt) %>% 
  get_summary_stats(c(TotalRRI, AphRRI, RhizRRI), type="mean_se")  %>%
  kbl(digits = 1) %>% kable_classic(full_width = F, html_font = "Arial")
```

**Now let's use the filter() function to look at only 2024 data**

```{r}
Assay %>% filter(Year == "2024") %>%
  group_by(Soil:Trmt) %>% 
  get_summary_stats(c(TotalRRI), type="mean_se")  %>%
  kbl(digits = 1) %>% kable_classic(full_width = F, html_font = "Arial")
```

**Let do only 2024 AND NOT the kabina treatment** 

```{r}
Assay %>% dplyr::filter(Year == "2024" & Trmt != "Kabina") %>% 
  group_by(Soil:Trmt) %>% 
  get_summary_stats(c(TotalRRI), type="mean_se")  %>%
  kbl(digits = 1) %>% kable_classic(full_width = F, html_font = "Arial")
```

**OR we can remove the kabina treatment from just the 2023 trial**; this is done by the OR operator.

```{r}
Assay %>% filter(Year == "2024" | Trmt == "Tachigaren") %>% 
  group_by(Soil:Trmt) %>% 
  get_summary_stats(c(TotalRRI), type="mean_se")  %>%
  kbl(digits = 1) %>% kable_classic(full_width = F, html_font = "Arial")
```


**REMINDER: Rather than creating new strings for everything, you can also pipe this directly into summary statistics!**

**Now lets put it all together to create another table**

```{r, warning=FALSE, message=FALSE}
Assay %>%
  group_by(Year:Soil:Trmt) %>% 
  get_summary_stats(c(TotalRRI, AphRRI, RhizRRI), type="common") %>% 
  
  dplyr::select(c('Year:Soil:Trmt', variable, mean)) %>%
  
  separate(col = 'Year:Soil:Trmt',
           into = c('Year', 'Soil', 'Trmt'),
           sep = ":") %>% 
  
  pivot_wider(
    names_from = variable,
    values_from = c(mean)) %>% 
  rename('Soil Location' = Soil,
         'Seed Treatment' = Trmt) %>% 
  
  filter(Year == "2024") %>% 

  kbl(digits = 1) %>%
    kable_classic(full_width = F, html_font = "Times New Roman") %>% 
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive"))
```

**We can even go futher...just bare with me!** 

```{r, warning=FALSE, message=FALSE}
Assay %>%
  filter(Year == "2024") %>% 
  group_by(Soil:Trmt) %>% 
  get_summary_stats(c(TotalRRI, AphRRI, RhizRRI), type="common") %>% 
  
  dplyr::select(c('Soil:Trmt', variable, mean, se)) %>% 
  
  mutate_at(vars(mean), funs(round(., 1))) %>% 
  # mutates the variable 'mean' by using the function to round the number to one digit after the decimal
  
  mutate_at(vars(se), funs(round(., 2))) %>% 
  # mutates the variable standard error
  
  mutate(se = paste0("(±", se, ")")) %>% 
  # takes the standard error column and pastes the plus/minus sign before the number and parethesis around the everything
  
  mutate(mean = paste(mean, se, sep = " "))  %>% 
  # take the new standard error column and pastes it after the mean column
  
  dplyr::select(-se) %>% 
  # drops se column
  
  separate(col = 'Soil:Trmt',
           into = c('Soil', 'Trmt'),
           sep = ":") %>% 
  
  pivot_wider(
    names_from = variable,
    values_from = c(mean)) %>% 
  rename('Soil Location' = Soil,
         'Seed Treatment' = Trmt) %>% 

  kbl() %>%
    kable_classic(full_width = F, html_font = "Times New Roman") %>% 
  kable_styling(bootstrap_options = c("hover", "condensed", "responsive"))
```